Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-23 analysis/paper/article.qmd
Last commit from JB! (I hope)

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. <!-- **MORE DETAILS ON THIS HERE—JB WILL ADD.** BM: ok, thank you! -->

@@

The desire to pursue an academic career also characterizes more recent PhD cohorts, including those within archaeology. Brami et al's survey found that despite the long odds and grueling precaraity, 71% of their early career respondents wished to remain in academia [-@bramiPrecariousFutureReflections2023, pp. 242]. Overall, the obsessive valorization of permanent faculty positions in the face of a precipitous decline in available jobs is indicative of the potent combination of denial and delusion inacross academia. As Cefkin and Schwegler argue "This asymmetry between the 'ideal' career path and the experiences of most graduates highlights the deep and growing chasm between the future that higher education institutions envision—and strongly incentivize—and what is actually happening" [@cefkin].

Returning to our results, the downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023]

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, subjective, and relational perspectives, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019]. This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

@@

Our results show that job ads offer a nuanced record of disciplinary history, documenting shifting trends in topics, methods, and geographic foci, with higher chronological resolution than many other sources, such as undergraduate textbooks [@lyman2010american]. The short periods of time spanning the rise and fall of some of the foci we observed have implications for prospective graduate students. A topic that is growing in popularity as a student begins their PhD may have peaked and be in decline before they graduate. Methods seem to have a much lower frequency of change in popularity relative to topical and geographic foci. One possible implication of this finding is that a graduate student who has invested in developing technical expertise in a method during their studies, in addition to a topic and region, might be less vulnerable to the vagaries of the job market than a student without a distinct area of technical expertise.

Jobs ads also demonstrate how archaeologists draw on current events and politics to prioritize topics, regions, and methods in their hiring plans. This result is consistent with prior findings, such @wilk1985ancient's analysis of publications that identified trends in the explanations of the Maya collapse during the 1960s-1970s that corresponded with major contemporary American political and social issues. This continual responsiveness to current events demonstrates that many archaeologists accept that our narratives of the past are entangled in the social and political milieu in which we work. Furthermore, it shows an active effort by archaeologists to take control of interpreting the past.

>>>>

A notable requirement that has emerged in recent years is a diversity statement, where the applicant describes their knowledge of, prior contributions to, and future goals for advancing diversity, equity, and inclusion. This requirement has been much-debated largely due to an experiment during 2016–2022 at several University of California campuses that used diversity statements as the first cut for selecting candidates for tenure-track faculty positions [@soucek2021diversity]. Only candidates that scored highly on their diversity statements would have the rest of their application evaluated. This experiment generated intense and widespread public debate about the merits and risks of requiring and using diversity statements. These debates drew attention to the challenges of hiring faculty from underrepresented minorities, and resulted in a variety of approaches to evaluating diversity, even leading some universities to entirely omit requirements for diversity statements in job applications [@guidenAreDiversityStatements]. Institutional attitudes to diversity are likely to remain in flux for the foreseeable future.The 2023 Supreme Court decision in Students for Fair Admissions v. Harvard, which effectively ended affirmative action at American colleges and universities, in tandem with the Trump administration’s 2025 targeting of DEI initiatives [@knox2025] foreshadow a shift in judicial and political currents that will reshape institutional approaches to—and even definitions of—diversity. The 180˚pivot in the past five years—from requiring carefully crafted diversity statements as a matter of course to sweeping policy shifts rendering such documents obsolete—indicates that national and institutional commitments to diversity may only be strategic plan deep.

@@

The desire to pursue an academic career also characterizes more recent PhD cohorts, including those within archaeology. Brami et al's survey found that despite the long odds and professional precarity, 71% of their early career respondents wished to remain in academia [= @bramiPrecariousFutureReflections2023, pp. 242]. Overall, the obsessive valorization of permanent faculty positions in the face of a precipitous decline in available jobs is indicative of the potent combination of denial and delusion across academia. As Cefkin and Schwegler argue "This asymmetry between the “ideal” career path and the experiences of most graduates highlights the deep and growing chasm between the future that higher education institutions envision—and strongly incentivize—and what is actually happening" [@cefkin].

Returning to our results, the downward trend in tenure-track positions during 2013-2019 may be related to declining undergraduate enrollment in anthropology since 2013 [@crambChangingProfileTenureTrack2022a]. The big dip during 2020-2021 is explained by the hiring freezes at many institutions resulting from the COVID-19 pandemic, which caused extreme disruption and uncertainty as universities focused on adapting to online instruction and assessment in an effort to minimize the spread of the virus [@woolstonJuniorResearchersHit2020]. A survey of early career researchers in archaeology captures the impact of this dip, with three-quarters of respondents experiencing negative impacts on their careers due to the pandemic [@bramiPrecariousFutureReflections2023].

@@

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. Over the course of the study period, we document a gradual decline in topics such as Complex societies and Archaeological science, the geographic foci of Mesoamerica and South America, and methods relating to Landscape archaeology. These trends are harder to explain as we cannot link their origins to a historical event such as the COVID-19 pandemic. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, subjective, and relational perspectives, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism [@hodderArchaeologicalProcessIntroduction1999; @johnson2019archaeological; @Fogelin2019].This theoretical trend might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for technological, functional, and compositional analyses. Other factors relevant to this decline may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past and an increased concern with climate change, environmental sustainability, and resilience, shifting attention away from the study of monumental architecture, elite societies, political hierarchies, and state systems.

@@

The data introduced in this paper were initially collected and analyzed in spring and summer 2024 and the manuscript was drafted in autumn 2024. Between submitting the manuscript and receiving our initial round of revisions in April 2025, the landscape of U.S. higher education has shifted dramatically. The first hundred days of the new administration focused on instilling a climate of “economic precarity and legal uncertainty” across the sector through a cthuluesque strategy that included withholding federal funding for colleges and universities, decimating the Department of Education, revoking international student visas, and attacking Diversity Equity, and Inclusion (DEI) initiatives as “unconstitutional” [@knox2025]. This “flood the zone” approach [@broadwater2025] shows no signs of abating; in April 2025, the president signed an executive order to overhaul the higher education accreditation system [@thewhitehouse2025], signifying the administration’s continued commitment to overhauling higher education remains intact.

Such challenges to the autonomy and self-governance in higher education are particularly remarkable given that many architects of the current administration hold degrees from the very institutions they are attempting to dismantle, from Donald Trump (Wharton School of Business), to J.D. Vance (Yale Law School), Chris Rufo (Georgetown), Stephen Miller (Duke University), and Peter Hegseth (Harvard Kennedy School). Reviewing the profiles of the board of trustees for the Heritage Foundation—the conservative think tank which generated the Project 2025 report that has been used as a blueprint for feeding US higher education “into a woodchipper” (Kunder 2025)—is also illustrative. Two thirds of the trustees on the Heritage Foundation board have advanced degrees, while one third have doctoral degrees from institutions such as Harvard, Oxford, the University of Colorado, and the University of Texas. Lindsey Burke, who authored the Project 2025 chapter on the Department of Education, which includes, holds a PhD from George Mason University. Such career trajectories are a testament to the elite chiasmus that decries the merits and value of higher education while leveraging the credentials granted by elite institutions for professional gain. As Ho notes of the parallel systems of credentialing that undergird the ideology of “smartness” amongst Wall Street investment banks, playing the role of “master of the universe” requires both “especially strong doses of self-confidence and institutional legitimation” [-@ho2009, pp. 41]. Even archaeology is susceptible to the power and pull of such prestige hierarchies, which influence everything from hiring networks [@speakman2018; @kawaSocialNetworkUS2019] to publishing decisions [@beck2021] to research design[@wobst1983].

To echo Paul Farmer, “all this is both interesting and horrible” [-@farmer2004, pp.316]. For the purposes of our paper, however, this sociopolitical tumult provokes a suite of new questions about academic futures. Many archaeologists now have urgent concerns about what topics are likely to be “safe” and lead to secure ongoing employment considering the current political and funding landscape. Will states with stronger cultural and environmental regulations offer more stability in employment, and if so, which states are likely to be the safest? These are legitimate and crucial questions, but our examination of the past decade of hiring dynamics in academic archaeology cannot predict the future, especially in a mercurial political and economic landscape that is experiencing tectonic shifts in bureaucracy, organization, and funding on a near-weekly basis. Our study does, however, point to at least four findings that have implications for the next ten years of hiring, despite the turbulent political climate of 2024–2025.

First, the economic uncertainty induced by the recent increases in tariffs is likely to have both short-term and long-term implications that slow economic growth [@schneider2025], effects which will seep into the higher education sector. The last two periods of major economic upheaval in the United States—the Great Recession and the COVID-19 pandemic—had pronounced and negative consequences for colleges and universities. These consequences were visible even within academic archaeology. Speakman et al., 2018, for example, show the precipitous decline in the proportion of PhDs hired into tenure-track anthropology departments after the last recession (see the steep decline in Figure 5 between 2007 and 2009), resulting in the “increased use of \[non-tenure track faculty\] by higher education institutions, and more anthropology doctorates entering the workforce”[@speakman2018, pp.12]. As our results demonstrate, during academic year 2020–2021, only 12 tenure-track jobs in archaeology were posted on the wiki, a 70% decrease relative to the mean of 39 per year posted during our survey period (Table 1). As Knox and Alonso [-@knox2025] highlight, hiring freezes and program cuts are already coming into effect just four months into the new administration. Given past patterns, it is likely that tariffs will result in tenure-track positions in academic archaeology becoming even more scarce, and faculty and students should prepare for this. Whether this signals a death knell for the discipline or an opportunity for re-envisioning and reconfiguring the field [@cefkin] is up for debate.

Second, our results show that hiring in academic archaeology is responsive to larger cultural and political shifts. The increased popularity of Indigenous and historical archaeology from 2019–2021 reflects the prominence and impact of movements such as Black Lives Matter and a then-growing interest in conversations about race and national history. The continual responsiveness to current events revealed by our analysis of archaeology job ads demonstrates that our narratives of the past are entangled in the social and political milieu in which we work. This finding is consistent with previous work—@trigger1984 outlined how archaeological research programs are shaped by national histories, while @wilk1985ancient, @blakey1983, @meltzer1983, and @soffer1983 have interrogated the entanglement of identity, ideology, national history, politics, and archaeology across space and time. Our study of job ads shows an active effort by archaeologists to take control of interpreting the past, offering an example of how disciplinary choices can shape archaeological research programs. That said, recent reports of National Science Foundation funding reviews triggered by a litany of keywords \[@johnson2025\]—including “cultural relevance,” “institutional,” “historically,” “socioeconomic,” and “systemic”—suggest the likelihood of impending topical shifts in archaeological research, as much an individual response to top-down funding pressures \[@wobst1983\] as a disciplinary response to the current historical moment. Third, the high chronological resolution of our analysis relative to previous studies \[@lyman2010american\] shows that the rise and fall of some of the foci we observed relatively span short periods of time. This limits the predictive power of our results and has implications for prospective graduate students and their mentors. A topic that is growing in popularity as a student begins their PhD may have peaked and be in decline before they graduate. Methods seem to have a much lower frequency of change in popularity relative to topical and geographic foci. One implication of this finding is that a graduate student who has invested in developing technical expertise in a method during their studies, in addition to a topic and region, might be less vulnerable to the vagaries of the job market than a student without a distinct area of technical expertise.

Finally, our work also reveals the market forces that shaped the early careers of the next generation of academic mentors. Assistant and associate professors who experienced the academic market as job seekers themselves between 2012–2023 probably anticipate that their students will need publications, external grants, teaching experience, interview preparation, extensive and tailored portfolios of job markets, and the ability to spend many years on the market before obtaining a permanent position, if they obtain one at all. This raises the question of whether mentors’ experiences on a market characterized by a recession and a pandemic will be enough to guide their students through the volatility of the next decade, and how to update our programs to prepare our graduates for success.
